Staged Abstract Interpreters

### 2019-09-07 paper/60_cases.tex
typo & revision

In this section, we show AC (closure generation) can be understood and
implemented as an instance of staging abstract interpreters. We first revisit
the original implementation of abstract compilation for 0-CFA, and then
reproduce their result by simply adding stage annotations.
The program generated by our approach provides approximately the same
improvement with regards to speed, but without changing a single line of the
analyzer program using type-based staging annotations. However, AC requires
more engineering effort, i.e., rewriting the whole analyzer into the
closure-generation form. Moreover, as shown in Section~\ref{staged_ds}, our
approach is able to not only remove the interpretive overhead, but also enable
several optimizations such as specialized data structures and equivalent
rewritings.

@@

\paragraph{Original AC}
The analysis presented by \citeauthor{Boucher:1996:ACN:647473.727587} is 0-CFA
for a small CPS language consisting of abstractions, applications, recursion, and
primitive operators. Figure \ref{compare_ac_sai} (left) shows the AC
implementation: different syntactic constructs are handled in different functions
(e.g., @compCall@, @compApp@, etc.).
Before refactoring into AC, these functions would take the store as a
regular argument; now every function returns a value of type @CompAnalysis@,
which is a function (closure) value that takes and returns stores. After the
first run of analysis, we have traversed the AST and obtained a single closure that
takes a store and returns a store. Therefore, the ASTs are static terms, and the
stores are dynamic terms. Now we can invoke the compiled closure with an initial
store to complete the analysis.

@@

On the other side, our approach works equivalently through staging: the
ASTs are static, whereas stores are dynamic; therefore, we annotate the type @Store@
with @Rep@ (shown in Figure \ref{compare_ac_sai} (right)). In fact, the only
changes of our approach using the LMS framework are the types of @Store@ to
@Rep[Store]@, since they will be known at the next stage: no other parts need to change.
The generated code
will consist entirely of looks-up and updating the store. The functions at the current stage are
eliminated in the residual program; therefore, we do not generate higher-order
functions, which provides additional performance improvement. Nevertheless,
the support from the LMS framework is necessary, such as staging support for @Map@.
We consider these efforts reusable and completely orthogonal to the
implementation of the analysis.

@@

In this section, to demonstrate that the SAI approach is widely applicable, we
extend and refactor the staged abstract interpreter we presented in Section
\ref{sai} to various control-flow analyses (CFA). We first discuss
calling-context sensitive analysis for staged analyzers by tweaking the
allocation strategy. Then we describe how to implement store-widening with
staged monads, which is a common optimization in CFA. Finally, we integrate
abstract garbage collection with staging. All of them are in fact orthogonal to
staging.

@@

The abstract addresses we introduced in Section \ref{sai} are simply identical
to the variable names, which is a context-insensitive schema.  Through tweaking
the address allocation strategy, we can achieve various context-sensitive
analyses \cite{DBLP:conf/icfp/Gilray0M16}. In this section, we demonstrate a
calling-context-sensitive analysis, i.e., $k$-CFA-like analysis
\cite{DBLP:journals/jfp/HornM12}, can be implemented our staged abstract
interpreter.
\begin{lstlisting}
  type Time = List[Expr]
  case class KCFAAddr(x: Ident, time: Time) extends Addr
\end{lstlisting}

>>>>

In this section, we show AC (closure generation) can be understood and
implemented as an instance of staging abstract interpreters. We first revisit
the original implementation of abstract compilation for 0-CFA, and then
reproduce their result by simply adding stage annotations.
The program generated by our approach provides approximately the same
improvement with regards to speed, but without changing a single line of the
analyzer program using type-based staging annotations. However, AC requires
more engineering effort, i.e., rewriting the whole analyzer into the
closure-generation form. Moreover, as shown in Section~\ref{staged_ds}, our
approach is able to not only remove the interpretive overhead, but also enable
several optimizations such as specialized data structures.

@@

\paragraph{Original AC}
The analysis presented by \citeauthor{Boucher:1996:ACN:647473.727587} is a 0-CFA
for a small CPS language consisting of abstractions, applications, recursion, and
primitive operators. Figure \ref{compare_ac_sai} (left) shows the AC
implementation: different syntactic constructs are handled in different functions
(e.g., @compCall@ and @compApp@).
Before refactoring to AC, the original functions take a store as a
regular argument. Now every function returns a value of type @CompAnalysis@,
which is a function value (i.e., closure) that takes and returns stores. After the
first run of the analysis, we have traversed the AST and obtained a single closure that
takes a store and returns a store. Therefore, the ASTs are static terms, and the
stores are dynamic terms. Now, we can invoke the compiled closure with an initial
store to complete the analysis.

@@

On the other side, our approach works equivalently through staging: the ASTs
are static, whereas stores are dynamic. Therefore, we annotate the type @Store@
with @Rep@ (Figure \ref{compare_ac_sai} (right)). In fact, the only changes of
our approach using the LMS framework are the types of @Store@ to @Rep[Store]@,
since they will be known at the next stage. There is no need for other changes.
The generated programs consist entirely of looks-up and updating operations
w.r.t the store.  The functions at the current stage are eliminated in the
residual program; therefore, we do not generate higher-order functions, which
provides additional performance improvement. Nevertheless, the LMS framework
provides support for necessary data structures, such as @Map@.  We consider
these efforts reusable and completely orthogonal to the implementation of the
analysis.

@@

In this section, to demonstrate that the SAI approach is widely applicable, we
extend and refactor the staged abstract interpreter we presented in Section
\ref{sai} to various control-flow analyses (CFA). We first discuss the
calling-context sensitive analysis for staged analyzers by tweaking the
allocation strategy. Then, we describe how to implement store-widening with the
staged monads, which is a common optimization in CFA. Finally, we integrate
abstract garbage collection with staging. All of them are orthogonal to
staging.

@@

The abstract addresses we introduced in Section \ref{sai} are simply identical
to the variable names, which is a context-insensitive schema.  Through tweaking
the address allocation strategy, we can achieve various context-sensitive
analyses \cite{DBLP:conf/icfp/Gilray0M16}. In this section, we demonstrate a
calling-context sensitive analysis, i.e., $k$-CFA-like analysis
\cite{DBLP:journals/jfp/HornM12}, can be implemented in our staged abstract
interpreter.
\begin{lstlisting}
  type Time = List[Expr]
  case class KCFAAddr(x: Ident, time: Time) extends Addr
\end{lstlisting}
